---
author: James  Watmough
date: "`r Sys.Date()`"
---

{{< include texmacros.qmd >}}

# Trigonometric Functions {#sec-trig-fun}

The trigonometric, or circular, functions find applications anywhere things oscillate.  Vibrating strings, water waves, seasonal temperatures or daily variations in light intensity are all examples where periodic functions arise.

## Assumed Background Knowledge

+ You should be familiar with most of the material in
[Openstax Volume 1 Section 1.3](https://openstax.org/books/calculus-volume-1/pages/1-3-trigonometric-functions)

  -  Convert angle measures between degrees and radians.
  -  Recognize the triangular and circular definitions of the basic trigonometric functions.
  -  Write the basic trigonometric identities.
  -  Identify the graphs and periods of the trigonometric functions.
  -  Describe the shift of a sine or cosine graph from the equation of the function.  

+ In particular, make sure you understand the connection of the 6 basic trig functions to a point on a unit circle and to triangles.  See [Figure 1.33](https://openstax.org/books/calculus-volume-1/pages/1-3-trigonometric-functions#CNX_Calc_Figure_01_03_007) in Openstax.

## Definitions of the sine and cosine

## Coordinates on a Circle

- Consider the relation $x^2+y^2=1$.

- We define the *cosine* and *sine* functions to be the $x$- and $y$-coordinates of the point as a function of the distance travelled counter-clockwise around the circle from the point (1,0).

:::: {.columns}

::: {.column width="60%"}

<!-- replace this with a high res png version using the standalone document style in latex -->

```{r engine='tikz', fig.ext="pdf"}
#| label: fig-trig-functions-circle
#| fig-cap: "Defining the sine and cosine functions using a circle"
#| echo: false

\begin{tikzpicture}[thick,scale=2,>=stealth]
  \draw[->] (-1.5,0) -- (1.5,0) node[below]{$x$};
  \draw[->] (0,-1.5) -- (0,1.5) node[left]{$y$};
  \draw (0,0) circle(1);
  \draw[fill] (.87,.5) circle(.02) node[right]{$(\cos(\theta),\sin(\theta))$};
  \draw (0,0) --(.87,.5);
  \draw[-stealth,thick] (1,0) arc(0:30:1);
  \draw (1,.3) node[left]{$\theta$};
\end{tikzpicture}

```

:::


::: {.column width="40%"}
[Desmos link](https://www.desmos.com/calculator/lmsgduhked)
:::

:::: 

## Ratios of sides of Right Triangles

<!-- replace this with a high res png version using the standalone document style in latex -->

```{r engine='tikz', fig.ext="pdf"}
#| label: fig-trig-functions-triangle
#| fig-cap: "Defining the sine and cosine functions using a triangle"
#| echo: false

\begin{tikzpicture}[scale=2]
  \draw (0,0) -- (2,0) node[below,pos=.5]{A} -- (2,1) node[right,pos=.5]{O} -- (0,0) node[above left,pos=.5]{H} ;
	\draw (.4,.1) node[right]{$\theta$};
\end{tikzpicture}

```

## Examples

### Sketching the Basic Trig Functions

::: {#exr-trig-sketch-1}

$y = 2 + 3\cos(2x-3)$

:::

::: {.callout-tip collapse=true title="solution"}

One aid to sketching functions is to evaluate the function at a few points.  The trick is picking the best points. Since we know the cosine is zero when its input is a multiple of $\pi$, we pick values for $x$ for which $2x-3$, the input to the cosine, is $\pi$.
That is, solve $2x-3 = \pi$ for $x$.  We find $x = \frac{1}{2}(3+\pi)$.  Of course this is only one multiple of $\pi$.  We said any multiple, so replace $\pi$ by $n\pi$ and let $n$ be any integer.  Similarly, we know the cosine is 1 if its input is $\frac{\pi}{2}$ plus any multiple of $\pi$.  Let's put this all in a table for easy reference.  We'll call the input to the cosine the angle $\theta$, pick values for $\theta$, and then compute $x = \frac{1}{2}(3+\theta)$ and $y = 2 + 3\cos \theta$.

|$\theta$         |$\cos \theta$ |$x$                          | $y$ |
|:---------------:|-------------:|:---------------------------:|----:|
|0                | 1            |$\frac{3}{2}$                | 5   |
|$\frac{\pi}{2}$  | 0            |$\frac{3}{2}+\frac{\pi}{4}$  | 2   |
|$\pi$            |-1            |$\frac{3}{2}+\frac{\pi}{2}$  | -1  |

:::

2. $y = -2\tan\big(\pi(x-2)\big)$

### Simplifying Trig Functions

1. $\sec x \sin x \cot x$

2. $\sec x - \cos x$

3. $\dfrac{1 + \tan^2 \alpha}{1 + \cot^2\alpha}$

4. $\dfrac{\sin x} {\cos x + 1} + \dfrac{\cos x - 1}{\sin x}$   

The last one is prob 150 in open stax. Add brackets to **remove those ambiguities!**





## Inverse Trigonometric Functions


The trigonometric functions and their inverses arise in a wide variety of problems involving triangles, lengths of curves and areas of surfaces.  Typical problems lead to triangles where the length of one of the sides is variable and one of the sides is fixed.  

A variety of notations are used for the inverse trig functions.  The inverse of the sine function is denoted by $\sin^{-1}$ or $\arcsin$.  In computer languages, the short form `asin` is often used.  Similar notations are used for the other standard trig functions.

::: {#exm-inverse-trig-1}

Suppose one end of a beam of unit length is lifted a height $x$ off the ground.  

- Find the horizontal distance between the ends of the beam.  
- Find the angle the beam makes with the ground.

Label the vertical of a right triangle as the dependent variable $x$ and the hypotenuse with the fixed unit length.

```{r engine='tikz', fig.ext="pdf"}
#| echo: false
#| height: 500

\begin{tikzpicture}[scale=5]
  \draw (0,0) -- (2,0) node[below,pos=.5]{$\sqrt{1-x^2}$} -- (2,1) node[right,pos=.5]{$x$} -- (0,0) node[above left,pos=.5]{1} ;
w (.4,.1) node[right]{$y$};
\end{tikzpicture}
```

$$x = \sin(y)$$
$$\sqrt{1-x^2} = \cos(y)$$
$$\tan y = \frac{x}{\sqrt{1-x^2}}$$

The horizontal distance between the ends of the beam is $\sqrt{1-x^2}$, based on Pythagoras' theorem. 
The angle $y$ is denoted by the inverse function $\sin^{-1}(x)$.
At this point we do not have an algorithm or rule for computing the inverse sine function, other than by use of a ruler and protractor.  We will learn methods for estimating values of this function towards the end of this course.

:::
     

::: {#exm-inverse-trig-2}

Suppose a boat is being pulled towards a dock.  Find the horizontal distance between the boat and the dock, the length of the rope played out and the angle the rope makes with the horizontal.

The rope is the hypotenuse of a triangle, and its length is variable.  We can assume the height of the rope at the dock is 1 unit.

```{r engine='tikz', fig.ext="pdf"}
#| echo: false

\begin{tikzpicture}[scale=2]
  \draw (0,0) -- (2,0) node[below,pos=.5]{$\sqrt{x^2-1}$} -- (2,1) node[right,pos=.5]{1} -- (0,0) node[above left,pos=.5]{$x$} ;
w (.4,.1) node[right]{$y$};
\end{tikzpicture}
```

$$x = \csc(y)$$
$$\sqrt{x^2-1} = \cot(y)$$

:::

::: {#exm-inverse-trig-2}

A third common scenario involves a balloon rising vertically.  Here the base of the triangle is the fixed length and the vertical is variable.

```{r engine='tikz', fig.ext="pdf"}
#| echo: false

\begin{tikzpicture}[scale=2]
	\draw (0,0) -- (2,0) node[below,pos=.5]{$1$} -- (2,1) node[right,pos=.5]{$x$} -- (0,0) node[above left,pos=.5]{$\sqrt{1+x^2}$} ;
	\draw (.4,.1) node[right]{$y$};
\end{tikzpicture}
```




$$x = \tan(y)$$
$$\sqrt{1+x^2} = \sec(y)$$
 
:::

::: {#exr-inverse-trig-1}

Use the above relations to verify the following strange identities.

$$x = \sin(\cos^{-1}(\sqrt{1-x^2})$$
$$x = \cot(\csc^{-1}(\sqrt{x^2-1})$$
$$x = \tan(\sec^{-1}(\sqrt{1+x^2})$$

Care must be taken in choosing the domain and range of the trig functions and their inverses.  These are always problem dependent.  Common choices for sine and cosine are shown in the graphs below. 

:::

```{r engine='tikz', fig.ext="pdf"}
#| echo: false

\begin{tikzpicture}
  \begin{scope}[scale=2]
    \draw[->] (-.2,0)--(3.3,0);
    \draw[->] (0,-1.2)--(0,1.2);
    \draw (0,1)--(-.1,1) node[left]{1};
    \draw (0,-1)--(-.1,-1) node[left]{-1};
    \draw (1.57,0)--(1.57,-.1) node[below]{$\frac{\pi}{2}$};
    \draw (3.14,0)--(3.14,-.1) node[below]{$\pi$};
    \draw[smooth,domain=0:3.14] plot (\x,{cos(\x r)});
    \draw (1.6,.5) node[right]{$y = \cos x$};
  \end{scope} 

  \begin{scope}[yshift=-2cm, xshift = 10cm,xscale=2.5,yscale=1.5]
    \draw[->] (0,-.2)--(0,3.3);
    \draw[->] (-1.2,0)--(1.2,0);
    \draw (1,0)--(1,-.1) node[below]{1};
    \draw (-1,0)--(-1,-.1) node[below]{-1};
    \draw (0,1.57)--(-.1,1.57) node[left]{$\frac{\pi}{2}$};
    \draw (0,3.14)--(-.1,3.14) node[left]{$\pi$};
    \draw[smooth,domain=0:3.14] plot ({cos(\x r)},\x);
    \draw (.5,1.6) node[right]{$y = \arccos x$};
  \end{scope} 

  \begin{scope}[yshift=-8cm, xshift=3cm, scale=2]
    \draw[->] (0,-1.2)--(0,1.2);
    \draw[->] (-1.6,0)--(1.6,0);
    \draw (0,1)--(-.1,1) node[left]{1};
    \draw (0,-1)--(-.1,-1) node[left]{-1};
    \draw (1.57,0)--(1.57,-.1) node[below]{$\frac{\pi}{2}$};
    \draw (-1.57,0)--(-1.57,-.1) node[left]{$-\frac{\pi}{2}$};
    \draw[smooth,domain=-1.57:1.57] plot (\x,{sin(\x r)});
    \draw (.5,1.6) node[right]{$y = \sin x$};
  \end{scope} 

  \begin{scope}[yshift=-8cm, xshift=10cm, xscale=2.5,yscale=1.5]
    \draw[->] (0,-1.8)--(0,1.8);
    \draw[->] (-1.2,0)--(1.2,0);
    \draw (1,0)--(1,-.1) node[below]{1};
    \draw (-1,0)--(-1,-.1) node[below]{-1};
    \draw (0,1.57)--(-.1,1.57) node[left]{$\frac{\pi}{2}$};
    \draw (0,-1.57)--(-.1,-1.57) node[left]{$-\frac{\pi}{2}$};
    \draw[smooth,domain=-1.56:1.56] plot ({sin(\x r)},\x);
    \draw (.5,1.6) node[right]{$y = \arcsin x$};
  \end{scope} 

\end{tikzpicture}
```
